%%
%% User Guide
%%
%% This file should be edited by user
%%

\chapter{User Guide}

You have implemented some system others will use? If you were in
their place what kind of documentation would you like to have in
order to start working?

For complex programs, a tutorial that guides the user through a
typical task showing screenshots of the intermediate states is
desirable.

\section{Types of Bachelor's Theses}

Of course, not all Bachelor's theses require a user guide.


\section{How to use the Thesis Template}

\subsection{Files to Edit}

You should edit the following files (unless you remove some of them,
see Section~\ref{section:removingchapters}):

\begin{itemize}
\item \tt title.tex
\item \tt abstract.tex
\item \tt acronyms.tex
\item \tt introduction.tex
\item \tt concepts.tex
\item \tt relatedwork.tex
\item \tt designapproach.tex
\item \tt implementation.tex
\item \tt results.tex
\item \tt conclusion.tex
\item \tt setupguide.tex
\item \tt userguide.tex
\end{itemize}

\subsection{Removing Chapters} \label{section:removingchapters}

Open the file \texttt{thesis.tex} and remove (or insert a comment)
at the lines with the include-commands, for example:

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|l|}
  \hline
  Before & After \\
  \hline
  % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
  \verb+\include{concepts}+ & \verb+\include{concepts}+ \\
  \verb+\cleardoublepage+ & \verb+\cleardoublepage+ \\
  \verb+\include{relatedwork}+ & \verb+%\include{relatedwork}+ \\
  \verb+\cleardoublepage+ & \verb+%\cleardoublepage+ \\
  \verb+\include{designapproach}+ & \verb+\include{designapproach}+ \\
  \verb+\cleardoublepage+ & \verb+\cleardoublepage+ \\
  \hline
\end{tabular}
\end{center}
  \caption{Removing the Chapter \texttt{Related Work}}\label{table:removing}
\end{table}

\subsection{Adding Chapters} \label{section:addingchapters}

Open the file \texttt{thesis.tex} and add an
\verb+\include{+\emph{filename}\verb+}+ command (followed by a
\verb+\cleardoublepage+) at the respective line. Then create a new
file \emph{filename}\verb+.tex+ in the same directory and write the
chapter's contents into it.

\subsection{Adding References} \label{section:addingreferences}

Whenever you \verb+\cite+ something, there must be a respective
entry in any of the included .bib files. To add such an entry, open
the respective bibfile (\eg \texttt{bibfile.bib}) with a text editor
and add the entry according to the bibfile syntax. The reference
will be added after running latex/bibtex/latex on your project.

\subsection{Removing References} \label{section:removingreferences}

Remove all \verb+\cite+ commands to this reference in the text and
the citation will not be listed in the bibliography after running
latex/bibtex/latex on your project. There is no need to remove the
reference from the .bib file.

\subsection{Changing the .bib Files to be used}

In the file \texttt{thesis.tex} there is a line with a
\verb+\bibliography+ command, that lists all used .bib files without
file extensions separated by commas.

\subsection{Troubleshooting}

If you make something wrong, you should get an error at compile
time, except for citation problems, like:

\begin{description}
\item[Adding or removing references does not work:]~~Perhaps there is
a spelling error in some .bib file, this causes the program
\texttt{bibtex} to abort execution leaving the old bibliography
unchanged.

\item[Changing a reference does not work:]~~Some systems only perform
a new bibtex run if there are missing references, delete the
\texttt{.aux} file in order to overcome this problem.

\item[References show a question mark:]~~Bibtex could not find a
matching entry in the bibfile for this reference, check the label
name.

\end{description}



%%
%% = eof =====================================================================
%%
